A Switching Planner for Combined Task and Observation Planning

### 2010-11-26 subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/dtpddl.tex
Few updates

Because PPDDL cannot model domains that feature partial observability,
we developed an extension of that language, called decision-theoretic
(DT)PDDL. This can express probabilistic models of the sensing
consequences of acting, to quantitatively capture unreliability in
perception. That expressive power is achieved by incorporating
perceptual analogues of fluent, predicate, and action definitions. In
detail, we have declarations of state characterising predicate and
fluent symbols according to the PPDDL syntax. In addition, we allow
two other declarations, for perceptual predicates and fluents
respectively. For example, suppose our robot is tasked with exploring
{\em locations} in order to identify the whereabouts of a {\em
visual-object}. We must describe state and perceptual facts that model
the {\em true}, resp. perceived, locations of objects. In DTPDDL,
these declarations appear as:

@@

The syntax and semantics for describing initial state distributions in
DTPDDL is taken verbatim from PPDDL. It is useful to present that
factored tree-like structure here, in order to aid us when discussing
our switching continual planner. The distribution is expressed in a
tree-like structure of terms. Each term is either: (1) atomic, e.g., a
state proposition such as $(=(\pp{is-in}~\pp{box})~\pp{office})$,
(2) probabilistic, e.g., $(\pp{probabilistic}~p_1 (T_1) .. p_n (T_n))$
where $T_i$ are conjunctive, or (3) a conjunct over probabilistic and
atomic terms. The root term is always conjunctive, and the leaves are
atomic. For example, the starting distribution for $\pp{DORA}$ might
be described thusly:\footnote{In PDDL, $(\pp{:init}~T_1..T_n)$
expresses the conjunctive root of the tree -- i.e., the root node
$(\pp{and}~T_1..T_n)$. Also, we shall write $\prop$, rather than
$(\pp{and}~\prop)$, it a conjunctive term contain a single atomic
subterm.}

>>>>

Because PPDDL cannot model domains that feature partial observability,
we developed an extension of that language, called Decision-Theoretic
(DT)PDDL. This can express probabilistic models of the sensing
consequences of acting, to quantitatively capture unreliability in
perception. That expressive power is achieved by incorporating
perceptual analogues of fluent, predicate, and action definitions. In
detail, we have declarations of state characterising predicate and
fluent symbols according to the PPDDL syntax. In addition, we allow
two other declarations, for perceptual predicates and fluents
respectively. For example, suppose our robot is tasked with exploring
{\em locations} in order to identify the whereabouts of a {\em
visual-object}. We must describe state and perceptual facts that model
the {\em true}, resp. perceived, locations of objects. In DTPDDL,
these declarations appear as:

@@

The syntax and semantics for describing initial state distributions in
DTPDDL is taken verbatim from PPDDL. It is useful to present that
factored tree-like structure here, in order to aid us when discussing
our switching continual planner. The distribution is expressed in a
tree-like structure of terms. Each term is either: (1) atomic, e.g., a
state proposition such as $(=(\pp{is-in}~\pp{box})~\pp{office})$,
(2) probabilistic, e.g., $(\pp{probabilistic}~p_1 (T_1) .. p_n (T_n))$
where $T_i$ are conjunctive, or (3) a conjunct over probabilistic and
atomic terms. The root term is always conjunctive, and the leaves are
atomic. For example, the starting distribution for $\pp{DORA}$ might
be described thusly:\footnote{In PDDL, $(\pp{:init}~T_1..T_n)$
expresses the conjunctive root of the tree -- i.e., the root node
$(\pp{and}~T_1..T_n)$. Also, we shall write $\prop$, rather than
$(\pp{and}~\prop)$, it a conjunctive term contains a single atomic
subterm.}

### 2010-11-26 subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/relatex.tex
Few updates

There have been a number of developments recently towards planning
under uncertainty using systems that were intended for sequential
planning in deterministic problems.  Notably,
\system{FFR$_a$}~\cite{yoon:etal:2007}, the winning entry from the
probabilistic track of the 2004 International Planning Competition.
In the continual paradigm, \system{FFR$_a$} uses the fast satisficing
procedure \system{FF}~\cite{hoffmann:nebel:2001} to compute sequential
plans and corresponding execution traces.
%%
More computationally expensive approaches in this vein combine
sampling strategies on valuations over {\em runtime variables} with
deterministic planning procedures. The outcome is typically a more
robust sequential plan~\cite{yoon:etal:2008}, or contingent
plan~\cite{majercik:2006}.
%%
Also leveraging deterministic planners in problems that feature
uncertainty, \system{Conformant-FF}~\cite{hoffmann:brafman:2006} and
$T_0$~\cite{palacios:geffner:2009} demonstrate how conformant planning
---i.e., sequential planning in unobservable worlds--- can be modelled
as a deterministic problem, and therefore solved using sequential
systems. In this conformant setting, advances have been towards
compact representations of beliefs amenable to existing best-first
search planning procedures, and lazy evaluations of beliefs. Most
recently this research thread has been extended to contingent planning
in fully observable non-deterministic
environments~\cite{albore:etal:2009}.

>>>>

There have been a number of developments recently towards planning
under uncertainty using systems that were intended for sequential
planning in deterministic problems.  Notably for example,
\system{FFR$_a$}~\cite{yoon:etal:2007}, the winning entry from the
probabilistic track of the 2004 International Planning Competition.
In the continual paradigm, \system{FFR$_a$} uses the fast satisficing
procedure \system{FF}~\cite{hoffmann:nebel:2001} to compute sequential
plans and corresponding execution traces.
%%
More computationally expensive approaches in this vein combine
sampling strategies on valuations over {\em runtime variables} with
deterministic planning procedures. The outcome is typically a more
robust sequential plan~\cite{yoon:etal:2008}, or contingent
plan~\cite{majercik:2006}. However, as we said in the introduction,
all these approaches struggle in partially observable domains as they
rely on being able to determine the state at all times.

Also leveraging deterministic planners in problems that feature
uncertainty, \system{Conformant-FF}~\cite{hoffmann:brafman:2006} and
$T_0$~\cite{palacios:geffner:2009} demonstrate how conformant planning
---i.e., sequential planning in unobservable worlds--- can be modelled
as a deterministic problem, and therefore solved using sequential
systems. In this conformant setting, advances have been towards
compact representations of beliefs amenable to existing best-first
search planning procedures, and lazy evaluations of beliefs. Most
recently this research thread has been extended to contingent planning
in fully observable non-deterministic
environments~\cite{albore:etal:2009}.
%%
The continual planning system that motivated our
project~\cite{brenner:nebel:jaamas09} also has this characteristic,
and has been applied in completely observable domains, particularly
those featuring multple communicating agents. The use of knowledge
operators in domains allows plans that act to gain knowledge, but the
approach assumes that such actions are deterministic and reliable, an
assumption that we relax in this work.

Our work is motivated by domains that contain a mixture of task
planning and observation planning. There have been a number of recent
papers representing observation planning problems as POMDPs and using
various techniques to manage the large state
space. \citeauthor{hippo-jnl}~(\citeyear{hippo-jnl}) take this
approach in a vision algorithm selection problem. In their case there
is a natural hierarchical decomposition of the problem which allows
them to solve large problems by breaking them into a set of small
POMDPs. \citeauthor{doshi08:pref_elic}~(\citeyear{doshi08:pref_elic})
represent a preference elicitation problem as a POMDP and take
advantage of symmetry in the belief space ---essentially the idea that
it does not matter what the value of the variable you are trying to
observe is--- to exponentially shrink the state space. Although we
have been actively exploring the \citeauthor{doshi08:pref_elic}
approach, we have yet to find those exploitable structures in our
domain due to the task planning component.

### 2010-11-26 subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/switching.tex
Few updates

We describe our continual planning approach that switches between
sequential and contingent sessions. Continual planning and execution
proceeds in a deterministic-action POMDP described in DTPDDL. During a
sequential session, planning is performed by a ``classical''
system,\footnote{That is, a system designed to solve fully observable
deterministic tasks.}  and execution proceeds according to the {\em
trace} computed by that system. Taking the form of a classical plan,
the {\em trace} specifies a sequence of POMDP actions that achieve the
agent's objectives in a deterministic approximation, i.e., {\em
determinisation}, of the problem at hand. More precisely, the trace is
an interleaved sequence of POMDP actions and {\em assumptive}
actions. The latter correspond to assumptions the planner makes about
the truth value of {\em runtime state variables} -- e.g. that a box of
cornflakes is located in the kitchen at the third plan step. Another
kind of assumption, called an {\em applicability} assumption, is made
when the trace includes an action $\action$ that is not applicable
with probability $1$ at the belief-state
\bstate\ that the system is projected to be in when \action\ is executed, 
i.e., $\exists\state\in\states\;
\bstate(\state) > 0$ and $\poss(\action)\not\subseteq\state$. By
scheduling
\action,  the serial planner makes an assumption about the
observability of the precondition $\poss(\action)$.

Our approach always begins with a sequential session. Non-assumptive
actions from the trace are executed until the applicability of the
next scheduled action, here \action, is questionable. A contingent
session then begins, which tailors sensory processing by the agent to
perceive whether the assumptions made in the trace hold. For each
active assumption in the trace, we add a {\em disconfirm} action to
the POMDP whose execution is rewarding (resp. costly) if the
assumption is false (resp. true). We also add one {\em confirm} action
that is rewarding (resp. costly) to execute if $\poss(\action)$ is
true (resp. false). If execution of the contingent plan applies a
disconfirm action, then a new sequential session begins at the current
underlying belief-state. If the confirm action is executed, the
sequential session is resumed, and \action\ is executed. Otherwise,
control rests in the continual session.
%%
Finally, because contingent planning is only practical in relatively
small POMDPs, contingent sessions proceed in an abstract decision
process determined by the current trace. That abstract process is
constructed by first omitting all variables from the model that are
not featured in the trace, and by then iteratively refining that
abstraction while the resulting model is of a practicable size.

PLEASE STICK A PARAGRAPH HERE, (SENSIBLY) REQUESTED BY MORITZ, ABOUT
HOW WE DO BELIEF REVISION ACROSS SWITCHING.

@@

For sequential planning the operators from the POMDP model are
available with semantics that accommodate the above abstraction as
follows. In the deterministic model a proposition $\prop$ can be
thought to have a ternary interpretation at a state \state, as either
{\em true}, written $\prop\in\state$, {\em false}, written
$\prop\not\in\state$, or {\em unspecified}, with a slight abuse of
notation written $\prop=\#\in\state$. For example, in $\pp{DORA}$
$(=(\pp{is-in}~\pp{box})\pp{office})\#\in\state_0$. For POMDP actions
$\action\in\actions$ in the deterministic model, if
$\prop \in \poss(\action)$, or if $\prop$ is the subject of a positive
or negative effect of $\action$, then $\action$ is not applicable
in \state\ if $\prop=\#\in\state$. Indeed, first an assumptive action
must be applied to make a truth assignment to \prop, and then that
truth assignment could be modified by POMDP operators. Lastly, we
address the switching semantics of action execution. If
$\action\in\actions$ has a precondition that is not true with
probability $1$ in the underlying belief $\bstate$, then before it is
executed blindly, the system switches to a contingent session,
discussed in detail in the following section.

@@

In a contingent session our switching planner solves a sensing problem
in an abstract processes defined in terms of the assumptive actions in
the current sequential plan, and the $(\pp{:init})$ declaration from
the original problem description.

@@

\noindent Where $\$(T_i)$ is a positive numeric quantity which
captures the reward the agent receives for correctly rejecting a
false assumption. The added dual actions are propositionally trivial
-- i.e., $\forall\state\in\states$ $\poss(\assumptiveDT{i})
\subseteq \state$ and $\assumptiveDT{i}(\state) \equiv \state$.  
%%
In order to reduce the scope of possible rejections, we compile the
serial trace into a partially ordered plan (POP). If $\action_i$ is
the sequentially scheduled execution that triggered the switch to the
contingent session, then we discard assumptive duals that are not
ordered before $\action_i$ in that POP.
%%
Also for the action $\action_i$, we include another propositionally
trivial action $\actions.\poss(\action_i)$ with the reward property:

@@

\noindent At this point, decision-theoretic planning would proceed in an
abstraction of the environment that does not contain a $\pp{cup}$,
where the applicability of POMDP actions is the same as we
defined it for the sequential session.

In a second step, we iteratively refine the above{\em relaxed}
declaration by adding terms from the original statement of
$(\pp{:init})$ while the resulting state space remains of a
practicable size. Of course, it is expensive to measure size exactly,
and in practise we estimate this to be $2^{|\propositions^\#|}$, where
$\propositions^\#$ is the set of propositions in the abstraction that
were not assigned the {\em unknown} value \#. Terms are proposed for
addition because they are probabilistic, and because they were omitted
in the {\em relaxed} visitation characterised by {\em assumptive}
actions in the sequential plan. In each iteration, we add one
candidate term with the largest number of parent terms already in the
candidate tree.

>>>>

We describe our continual planning approach that switches between
sequential and contingent sessions. This proceeds by interleaved
planning and execution in a deterministic-action POMDP described in
DTPDDL. During a sequential session, planning is performed by a
``classical'' system,\footnote{That is, a system designed to solve
fully observable deterministic tasks.}  and execution proceeds
according to the {\em trace} computed by that system. Taking the form
of a classical plan, the trace specifies a sequence of POMDP
actions that achieve the agent's objectives in a deterministic
approximation, i.e., {\em determinisation}, of the problem at
hand. More precisely, the trace is an interleaved sequence of POMDP
actions and {\em assumptive} actions. The latter correspond to
assumptions the planner makes about the truth value of {\em runtime
state variables} -- e.g. that a box of cornflakes is located in the
kitchen at the third plan step. Another kind of assumption, called an
{\em applicability} assumption, is made when the trace includes an
action $\action$ that is not applicable with probability $1$ at the
belief-state
\bstate\ that the system is projected to be in when \action\ is
scheduled for execution, i.e., $\exists\state\in\states\;
\bstate(\state) > 0$ and $\poss(\action)\not\subseteq\state$. By
scheduling
\action,  the serial planner makes an assumption about the
observability of the precondition $\poss(\action)$.

Our approach always begins with a sequential session. Non-assumptive
actions from the trace are executed until the applicability of the
next scheduled action is uncertain.  A contingent session then begins,
which tailors sensory processing by the agent to determine whether the
assumptions made in the trace hold. For each {\em active} assumption
in the trace, we add a {\em disconfirm} action to the POMDP whose
execution is rewarding (resp. costly) if the assumption is false
(resp. true). We also add one {\em confirm} action that is rewarding
(resp. costly) to execute if $\poss(\action)$ is true
(resp. false). If execution of the contingent plan applies a
disconfirm action, then a new sequential session begins at the current
underlying belief-state. If the confirm action is executed, the
sequential session is resumed, and
\switchAction\ is executed. Otherwise, control rests in the continual
session.

Because contingent planning is only practical in relatively small
POMDPs, contingent sessions proceed in an abstract decision process
determined by the current trace. That abstract process is constructed
by first omitting all variables from the model that are not featured
in the trace, and by then iteratively refining that abstraction while
the resulting model is of a practicable size.

Finally, whether proceeding in a sequential or contingent session, our
continual planner maintains a factored representation of successive
{\em belief-states}, by performing belief revision according to
Eq~\ref{eq:revision}. Moreover, our internal representation of that
distribution corresponds more-or-less directly to an $(\pp{:init})$
declaration in DTPDDL. Our approach uses that {\em underlying}
belief-state as: (1) the source of determinisations for sequential
planning, (2) the basis for switching, and (3) the source for state
abstraction in contingent sessions.

%% PLEASE STICK A PARAGRAPH HERE, (SENSIBLY) REQUESTED BY MORITZ, ABOUT
%% HOW WE DO BELIEF REVISION ACROSS SWITCHING.

@@

For sequential planning the operators from the POMDP model are
available with semantics that accommodate the above abstraction as
follows. In the deterministic model a proposition $\prop$ can be
thought to have a ternary interpretation at a state \state, as either
{\em true}, written $\prop\in\state$, {\em false}, written
$\prop\not\in\state$, or {\em unspecified}, with a slight abuse of
notation written $\prop\#\in\state$. For example, in $\pp{DORA}$
$(=(\pp{is-in}~\pp{box})\pp{office})\#\in\state_0$. For POMDP actions
$\action\in\actions$ in the deterministic model, if
$\prop \in \poss(\action)$, or if $\prop$ is the subject of a positive
or negative effect of $\action$, then $\action$ is not applicable
in \state\ if $\prop\#\in\state$. Indeed, first an assumptive action
must be applied to make a truth assignment to \prop, and then that
truth assignment could be modified by POMDP operators.

Addressing now the switching semantics of action execution. If
$\action\in\actions$ has a precondition that is not true with
probability $1$ in the underlying belief $\bstate$, then before it is
executed blindly, the system switches to a contingent session,
discussed in the following section. In detail, we halt the sequential
session at \action\ if $\exists
\state$ s.t. $\bstate(\state)>0$, 
$\prop\in\poss(\action)$ , and $\prop\not\in\state$. According to the
semantics of action execution there must be a $\state'$ s.t.
$\bstate(\state')>0$ and $\prop\in\state'$. Otherwise, \action\ must
not have been scheduled at \bstate.

then $\reward(\state,\action)=10$, and
otherwise if $\prop\not\in\state'$ then
$\reward(\state',\action)=-10$.

@@

In a contingent session our switching planner solves a sensing problem
in an abstract processes defined in terms of the assumptive actions in
the current trace, and the $(\pp{:init})$ declaration from the
original problem description.

@@

\noindent Where $\$(T_i)$ is a positive numeric quantity which
captures the reward the agent receives for correctly rejecting a
false assumption. The added dual actions are propositionally trivial
-- i.e., $\forall\state\in\states$ $\poss(\assumptiveDT{i})
\subseteq \state$ and $\assumptiveDT{i}(\state) \equiv \state$.  
%%
In order to reduce the number of possible rejections we do not
consider assumptions that are not {\em active} with respect to the
action, $\action_i$, whose scheduled execution switch the system into
the contingent session. A assumption $\assumptiveS{i}$ is active if
$\action_i$ is not applicable in the sequential model unless
$\assumptiveS{i}$ is scheduled in the trace prefix to $\action_i$. For
example, if $\pp{DORA}$'s trace is:

\[
\begin{array}{l}
\actions^{\circ}(.8;(=(\pp{is-in}~\pp{box})\pp{office}));\\
\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}));\\
(\pp{look-for-object}~\pp{DORA}~\pp{box}~\pp{office});\\
(\pp{look-for-object}~\pp{DORA}~\pp{cup}~\pp{kitchen});\\
(\pp{report-is-in}~\pp{box}~\pp{office}); \\
(\pp{report-is-in}~\pp{cup}~\pp{kitchen})
\end{array}
\]

\noindent Taking the switching action $\action_i$ to be
$(\pp{look-for-object}~\pp{DORA}~\pp{box}~\pp{office})$, we have that
$\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ is not
active, and therefore exclude
$\actions^{\bullet}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ from the
POMDP posed to the contingent session.

Again taking $\action_i$ to be the switching action, we include
another propositionally trivial action $\actions.\poss(\action_i)$
with the reward property:

@@

\noindent At this point, the session would proceed in an
abstraction of the environment that does not contain a
$\pp{cup}$. With regards to abstraction, the contingent and serial
sessions apply the same applicability conditions ---i.e., that each
effect and condition $\prop$ is specified $\prop\in\state$
(resp. $\prop\#\in\state$)--- to actions from the DTPDDL problem
description.

In a second step, we iteratively refine the relaxed declaration by
adding terms from the original statement of $(\pp{:init})$ while the
number of abstract states that occur with non-zero probability
according to that refined declaration remains of a practicable
size. In detail, for each excluded atomic term, we compute the {\em
entropy} of the corresponding proposition, {\em conditional} on the
active assumptions. The {\em conditional entropy} measure
is higher the more uncertain that proposition is given truth
assignments to propositions that are the subject of active
assumptions. Propositions that in the first place were excluded are
iteratively added to the problem posed to the conditional session in
increasing order according to that measure.

%% In particular, we compute the {\em conditional entropy} of
%% each excluded proposition given the active assumptions. Propositions
%% are candidates for addition in increasing order according to that
%% measure.

%% Of course, it is expensive to measure size exactly,
%% and in practise we estimate this to be $2^{|\propositions^\#|}$, where
%% $\propositions^\#$ is the set of propositions in the abstraction that
%% were not assigned the {\em unknown} value \#. Terms are proposed for
%% addition because they are probabilistic, and because they were omitted
%% in the {\em relaxed} visitation characterised by {\em assumptive}
%% actions in the sequential plan. In each iteration, we add one
%% candidate term with the largest number of parent terms already in the
%% candidate tree.
